---
title: "compstatsR session 3:<br>C++ and Rcpp"
author: "Ben Bolker"
date: "08 November 2016"
bibliography: compstatsR.bib
output:
   ioslides_presentation
---
<style>
.refs {
   font-size: 12px;
}
h2 { 
 color: #3399ff;		
}
h3 { 
 color: #3399ff;		
}
.title-slide {
   background-color: #55bbff;
}
</style>
<!-- Limit image width and height -->
<!-- 
img {     
  max-height: 560px;     
  max-width: 800px; 
}
-->
<style type="text/css">
img {     
  max-height: 500px;     
  max-width: 720px; 
}
</style>
<!-- n.b. comment must go outside style tags -->
<!-- https://css-tricks.com/snippets/css/simple-and-nice-blockquote-styling/ -->
<!-- quotes were "\201C""\201D""\2018""\2019"; -->
<style>
blockquote {
  background: #f9f9f9;
  border-left: 10px solid #ccc;
  margin: 1.5em 10px;
  padding: 0.5em 10px;
  quotes: "\201C""\201C""\201C""\201C";
}
blockquote:before {
  color: #ccc;
  content: open-quote;
  font-size: 4em;
  line-height: 0.1em;
  margin-right: 0.25em;
  vertical-align: -0.4em;
}
blockquote p {
  display: inline;
}
</style>
<style>
div#before-column p.forceBreak {
	break-before: column;
}
div#after-column p.forceBreak {
	break-after: column;
}
</style>

```{r setup, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
insert_fun <- function(name) {
    read_chunk(lines = capture.output(dump(name, '')),
               labels = paste(name, 'source', sep = '-'))
}
```
# When to use C++ 

## Don't (from [Rich Fitzjohn][1])

- Slower development
- Have to learn/remember two languages
- Profiling is harder
- Debugging is harder
- Platform independence is harder
- Distributing code is harder

[1]: https://github.com/richfitz/rc2016/blob/master/slides.md "R & C / C++"

## Instead

- optimize and profile
- use existing library code
- check algorithms  
(RF: "Using C will not magically make a bad algorithm faster")

## Why *should* you?

- Proven need for speed
- Very large objects: avoid copying
- Linking to external software libraries

> Typical bottlenecks that C++ can address include:
> 
> - Loops that can’t be easily vectorised because subsequent iterations depend on previous ones.
> - Recursive functions, or problems which involve calling functions millions of times. The overhead of calling a function in C++ is much lower than that in R.
> - Problems that require advanced data structures and algorithms that R doesn’t provide ...

# C++ super-basics

## C++ vs. R

C++ is:

- older
- [much more widely used][2]
- lower-level ("close to the metal")
- strongly object-oriented ...
- superset of (simpler) C
- compiled vs. interpreted

[2]: http://spectrum.ieee.org/computing/software/the-2016-top-programming-languages "top programming languages"

## C++ differences from R

- syntax
    - statements end with semicolons
- zero-indexing
- stronger/static typing
    - variable declaration
	- scalars and vectors are different


## example (modified from HW) {.columns-2}

```{r rcpp1,message=FALSE,cache=TRUE}
library(Rcpp)
cppFunction(
 'int subtract(int x, int y) {
    int diff = x - y;
    return diff;
  }')
subtract(4,5)
subtract(y=4,x=5)
```

- all explicitly typed
(return value, args, temp variable:
all integer *scalars*)
- explicit `return` statement
- `cppFunction` magically finds function name
- named arguments   
(no default values)

## Switch to *Advanced R* examples

([here](http://adv-r.had.co.nz/Rcpp.html#rcpp-intro))

## Summary

- `for` loops
    - `for (<init>; <test>; <increment>)`, e.g.
    - `for (i=0; i++; i<N)`
- scalar types: `double`, `int`, `String`, `bool`
- vector types: `NumericVector`, `IntegerVector`, `CharacterVector`, `LogicalVector`  
(R `length()` == C++ `.size()`)
- matrix types: `NumericMatrix`, etc.  
(index via `m(i,j)`; `.nrow()`, `.ncol()`)

## Connecting C++ to R

- via `cppFunction` (quick and dirty/magic)
- via `sourceCpp` (better)
- via an R package (best for long term)

## using sourceCpp

- C++ code in a separate file
- enables syntax highlighting
- need

```
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]  // for each function
```

`sourceCpp("path/filename.cpp")`

## Output from Rcpp

- essential for debugging!
- use either `Rprintf` (R-style) or `Rcout` (C++-style) printing:

```{r print}
cppFunction('void print_stuff1(double val) {
   Rprintf("hello world: %f\\n",val);
}')
print_stuff1(pi)
cppFunction('void print_stuff2(double val) {
   Rcout << "hello world " << val << std::endl;
}')
print_stuff1(pi)
print_stuff2(pi)
```

## Gibbs sampling example

From [Darren Wilkinson's blog](https://darrenjw.wordpress.com/2011/07/16/gibbs-sampler-in-various-languages-revisited/):

$$
\begin{split}
f(x,y) & = kx^2 \exp(-xy^2-y^2+2y-4x) \quad x >0, y \in {\mathbf R} \\
x |y & \sim \textrm{Gamma}(3,y^2+4) \\
y |x & \sim \textrm{Normal}\left(\frac{1}{1+x},\frac{1}{2(1+x)}\right)
\end{split}
$$

([back to *Advanced R*](http://adv-r.had.co.nz/Rcpp.html#rcpp-case-studies))

## Sugar

- [syntactic sugar](https://en.wikipedia.org/wiki/Syntactic_sugar)

>  If there’s a sugar version of the function you’re interested in, you should use it: it’ll be both expressive and well tested. Sugar functions aren’t always faster than a handwritten equivalent, but they will get faster in the future as more time is spent on optimising Rcpp. ([AdvR](http://adv-r.had.co.nz/Rcpp.html#rcpp-case-studies#sugar))

- automatic vectorized arithmetic, logical comparisons, math functions, probability distributions ...

## Sugar example

```{r sugar,cache=TRUE}
cppFunction(
"NumericVector pdistC2(double x, NumericVector ys) {
  return sqrt(pow((x - ys), 2));
}
")
pdistC2(2,-2:4)
```

## libraries

- `RcppEigen` (Eigen library)
- `RcppArmadillo` ([Armadillo](http://arma.sourceforge.net/) library)

Super-simple example. 

```{r arma}
library(RcppArmadillo)
cppFunction(depends="RcppArmadillo",
    "arma::mat matMult(arma::mat x, arma::mat y) {
       return (x%y);
}
")
matMult(matrix(1:4,2),matrix(1:4,2))
```

## More on Armadillo

- automatic conversion from R matrix $\leftrightarrow$ `NumericMatrix` $\leftrightarrow$ `arma::mat` (elsewhere may need `as<arma::mat>` to convert)
- `.fill` to initialize matrices
- syntax generally MATLAB-like (!!): [table](http://arma.sourceforge.net/docs.html#syntax)

## Resources

- books [@eddelbuettel_seamless_2013;@wickham_advanced_2014]
- [Rcpp on the author's web site](http://dirk.eddelbuettel.com/code/rcpp.html)
- [quick reference guide](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-quickref.pdf) (PDF)
- [Rcpp gallery](http://gallery.rcpp.org/)
   - [stochastic SIR model](http://gallery.rcpp.org/articles/epidemiological-compartment-model/)
   - [EM algorithm for probit regression](http://gallery.rcpp.org/articles/EM-algorithm-example/) (with Armadillo)
- [another basic intro to RcppArmadillo](http://q-aps.princeton.edu/sites/default/files/q-aps/files/slides_day4_am.pdf)

<!-- http://stackoverflow.com/questions/10474107/fast-bounding-of-data-in-r/10475036#10475036 -->

## References {.refs}

